\documentclass{article}

\usepackage{graphicx}
\usepackage{minted}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage[margin=1.0in]{geometry}
\usepackage{multicol}
\usepackage{titlesec}
\usepackage{filecontents}
\usepackage{url}

\definecolor{airforceblue}{rgb}{0.36, 0.54, 0.66}
\definecolor{secblue}{rgb}{0.18, 0.27, 0.33}

\input{style.tex}

\pagestyle{plain}

\title{Hiss: Static Duck Type Checking for Python\\ (Proposal)}
\author{Joshua Rahm}

\begin{filecontents}{bib.bib}
@misc{logic,
  author = {Ravi Chugh, Patrick M. Rondon, Ranjit Jhal},
  title = {{Nested Refinements: A Logic for Duck Typing}},
  howpublished = "\url{http://people.cs.uchicago.edu/~rchugh/static/papers/popl12-nested.pdf}",
  note = "[Online; accessed 10-November-2015]"
}

@misc{nst,
  title = {{Nomitive and Structural Typing}},
  howpublished = "\url{http://c2.com/cgi/wiki?NominativeAndStructuralTyping}",
  note = "[Online; accessed 13-November-2015]"
}

@misc{dynamic,
  author = {Jonathan Eifrig, Scott Smith, Valery Trifonov},
  title = {{Type Inference for Recursively Constrained Types and its Application to OOP}},
  howpublished = "\url{http://citeseerx.ist.psu.edu/viewdoc/download;jsessionid=F4F3BAF0FC6AB2A0681AA4428CFD4C4A?doi=10.1.1.92.6519&rep=rep1&type=pdf}",
  note = "[Online; accessed 13-November-2015]"
}
\end{filecontents}


\begin{document}
\nocite{*}
\maketitle
\hrule

% \raggedcolumns
\begin{multicols}{2}
% \raggedcolumns

\begin{abstract}

    In recent years, dynamically typed languages, specifically Python, have
    become extremely popular, especially for the object-oriented, imperative
    paradigms. Like most technologies, there are benefits and there are cons.
    Some of the benefits include generally less typing and quicker development
    time. No matter how quick the new development time, One of the largest
    drawbacks of dynamic languages is the lack of good static checking,
    specifically type checking. The fact is that while dynamic languages give
    the programmer huge flexibility in terms of type-introspection at runtime.
    Most programmers, however, choose not to use these features and instead
    treat Python as if it was statically typed and just keep track of the types
    in their heads while programming. For programs like these, it should be
    able to infer types of variables and detect inconsistencies and report to
    the programmer.

\end{abstract}

\section*{Problem}

The problem facing many programmers is a dilemma of choosing between the
flexibility and the terseness of dynamically typed languages such as Python and
the safety of statically typed languages. Indeed, dynamic typing the past
several years has gained some immense popularity among developers for its
relative ease; however, the lack of good static checking can make it very
cumbersome to hunt down and eliminate even simple bugs which a compiler
would most likely catch in a statically typed language.

The real crime is the fact that most programs, regardless of typing discipline,
are written in a way that is intended to by type-safe.  In addition, most of
the code written even conforms to type erasure semantics, meaning there is no
type-introspection being done at runtime. Under these conditions, it should be
easy for an automated reasoning engine to detect inconsistencies in code intended
to fit this definition of safety.

Such a tool would cut development time even more than these dynamic languages
have already. The major hurdle when it comes to creating such a tool is
the lack of explicit typing in nearly all aspects of a program. This makes is
more difficult as it prohibits the tracking of strict types, which means
it must implement forms of type inference for structural and duck typing
disciplines.

\section*{Proposal}

I propose that as an implementation project, I will implement a subset of the
static checker described in the previous project for a subset of the Python
programming language. The exact subset has yet to be defined, but it will
operate under sanely typed Python programs, meaning programs which are intended
to be type-safe and adhere to type-erasure semantics.

The more exact terms of the implementation are described in later sections.

The project name is Hiss, after (of course) the sound a snake makes.

\section*{Duck Typing}

Duck typing, specifically the ability to infer types in a Duck-Typed system,
is at the heart of the implementation of Hiss.

A misconception is that Duck Typing is only associated with dynamically typed
languages. This is false, although it is true that most modern dynamically typed
languages implement a form of duck typing, it is not rare to see a statically
typed language implement some duck typing. A very common example is the following
example using C++ templates.

\inputminted{cpp}{test1.cpp}

This code snippet will compile, and will behave as expected. C++ is able
to infer types properly, statically check them, and compile. Even though
\emph{Person} and \emph{Duck} are not a part of the same inheritance structure,
they both may still be used in the \emph{InTheForest} function because they
happen to have the correct functions.

If C++ can compile this, then the python analog should be checkable:

\inputminted{python}{test1.py}

Alas, there are very few, if any, static verifiers out there that can actually
check this code. In fact, changing the definition of \emph{InTheForest}
to be:

\begin{minted}{python}
def InTheForest(duck):
    duck.quack()
    duck.walk()
    duck.feathers()
\end{minted}

there is absolutely no indication from pylint (a common python lint checker) that
this will blatantly fail at runtime because a \emph{Person} does not have the
method \emph{feathers}. Yet a similar change in the C++ code will result in a
very quick compile error.

\subsection*{Duck Type Inference \& Structural Typing}

As stated previously, at the heart of Hiss is the need to be able to use logic
to infer the types of a Duck-Typed programming language. This requires a very
different approach to typing than traditional type systems like those in Java
or C++. What is required is a model of types which is not based on the name of
the type, but rather a model of typing based on the attributes of the values.
This a concept that Structural typing attempts to formalize. In a structural
type system, for an object $O$ to be compatible with a type $T$, $O$ must implement
a superset of attributes defined in $T$.\cite{nst}

\subsection*{Constrained Typing with Type Inference}

More languages these days like Haskell are implementing a typing discipline
knows as constrained types. These types allow the programmer to define function
and attributes to a \emph{class of data types}. For example, a ``Readable'' type
may have a function called \emph{read}, likewise a ``Monoid'' type may have a
function called ``zero'' and a function called ``append''. Readable and Monoid are
example of constraints on a type. When performing type-inference on them, they
help to provide the most generic, but complete constraint for a type. So if a
type $\alpha$ is observed to be used in an ``append'' call, it can be inferred
that $\alpha$ is \emph{constrained} by the class Monoid.

\subsection*{Inference}

When witnessing an assignment to a variable $v$, we can extrapolate the
expected \emph{structural type} of $v$. From there, as Hiss linearly traces
through the program, every method invoked on $v$ and every attribute accessed
on $v$ can be verified to be a part of $v$'s structural type. If not, a warning
will be raised to notify the programmer that a potential type error has
occurred.

In the case that an assignment to $v$ is a base case, meaning the assignment
is from either a string literal, a number, or a constructor, then the full
type as well as the structural type is then known by Hiss and further inferences
may be made.

\subsubsection*{Functions}

Structural types may be given to a function's return type and argument types by
this inference of the structural types of the arguments as well as the return
type. This can be used both to infer types of the parameters and the return type
so that all invocations of the function may also be inferred with the correct
structural types.

\subsubsection*{Any Type}

To completely express this type system, there must be a notion of the "any type"
meaning that there are no clues to the actual type of a variable and therefore
its structural type is a wild card of sorts and cannot be determined. The Any Type
will come into play when using an opaque library and it is not possible to determine

\section*{Timeline}

As the weeks progress, I will work on the implementation of this project. The
implementation will be written in Haskell since there is a fairly mature and reliable
Python parser for the language. After the first week, I plan to have basic scaffolding
done, then over the break and the following week I plan to implement some of the
basic functionality. Finally, the last week of the semester will be dedicated into
formalizing the results and putting together a short writeup about the findings.

\bibliographystyle{abbrv}
\bibliography{bib}

\vspace{\textheight}
\end{multicols}
\end{document}
